On the robustness and consistency of phylogeographic inference

### 2024-06-21 manuscript/phylogeo.tex
edit manuscript

\maketitle
As organisms replicate, they also disperse in space, resulting in mixing of the population and exploration of new habitats.
The reconstruction of ancestral locations and past migrations from sampling locations of extant individuals, typically in combination with genome sequence information to infer the phylogeny, is known as phylogeography.
Phylogeographic methods are implemented in popular evolutionary analysis software such as BEAST \citep{lemey_bayesian_2009,lemey_phylogeography_2010}.
The underlying model of dispersal is typically assumed to be diffusive, that is the probability of sampling a descendant at position $\rvec_c$ after time $t$ when the parent was at position $\rvec_p$ is given by
\begin{equation}
    P(\rvec_c| \rvec_p, t) = \frac{1}{4\pi D t}e^{\frac{|r_c - r_p|^2}{4Dt}}
\end{equation}
where $D$ is a diffusion constant with dimensions $\mathrm{length}^2/\mathrm{time}$, and we have assumed a two-dimensional space and ignored anisotropic dispersal.
Such a diffusive process is the simplest and most natural choice in absence of directed motion or long range migration.
It is also mathematically convenient as unknown ancestral locations can be integrated in closed form and the marginal distribution of each position can be calculated exactly for a fixed tree along with the likelihood.
The total likelihood of the configuration is then assumed to be the product of the spatial likelihood and the phylogenetic likelihood, that is the tree topology and branching rates of the tree are assumed to be independent of the spatial process.
This independence assumption is related to the common assumption that all individuals in the population are equally fit.

However, population dynamics is often tightly coupled to spatial location.
Environmental change could mean that a habitat of a population is shifting, resulting in population growth and rapid branching at the expanding edge of the habitat and/or contraction in other areas.
Similarly, invasive species or pathogen might spread into previously areas that support high population densities but where previously empty.
Examples of such shifting habitats are common and range from geological time scales, over decades (climate change), to months (seasonal fluctuations).
In all such cases, the replication process determining the phylogeny of the sample and the spatial location are strongly coupled.

Such coupling can be accommodated in structured coalescent models where the population is divided into discrete subpopulations that exchange migrants and have different population dynamics \citep{vaughan_efficient_2014}.
But the extent to which violations of the independence assumption affect continuous phylogeographic inferences is not well understood.
To accommodate variation not captured by the simple diffusion model, ``relaxed random walk'' models allow the diffusion constant $D$ to vary across the tree (the diffusion constant is integrated over a prior distribution) \citep{dellicour_relax_2021,lemey_phylogeography_2010}.

\begin{figure*}[tb]
    \includegraphics*[width=0.7\textwidth]{figures/illustration_tree.pdf}
    \caption{\label{fig:illustration_tree}{\bf Illustration Yule and Kingman trees and the spatial location of the tips.}
    The total tree length of Yule trees is proportional to the number of tips $n$ and the time to the MRCA is $\log(n)$.
    Kingman trees have a total tree length that is proportional to $\log(n)$ and are characterized by rapid merging close to the present. }
\end{figure*}

In a typically Bayesian phylogeographic inference problem, the input data are typically viral genome sequences as well as their sampling times and sampling locations.
Software like BEAST will then sample the posterior distribution of the phylogenetic tree topology, the sampling times and locations of ancestral nodes in the tree, and model parameters like the diffusion constant $D$ and the evolutionary rate $\mu$.
Results can be estimates of the ancestral location of a particular groups of samples, or summary statistics of the dispersal process.
Popular summaries include empirical estimates of the diffusion constant  \citep{pybus_unifying_2012,trovao_bayesian_2015} and a so-called \emph{dispersal velocity} \citep{dellicour_using_2017}. These are calculated from the estimated displacements along branch $i$ with parent $p_i$ given by $d_i = |\rvec_{p_i} - \rvec_{i}|$ and the elapsed time $\Delta t_i$ as
\begin{equation}
    \label{eq:dispersal_parameters}
    D_w = \frac{\sum_{i\in B}d_i^2}{4\sum_{i\in B} \Delta t_i} \quad \mathrm{and}  \quad v_w = \frac{\sum_{i\in B} d_i}{\sum_{i\in B} \Delta t_i}
\end{equation}
where $B$ is the set of all branches of the tree.
Effectively $D_w$ divides the sum of all observed squared displacements by the total time elapsed on the tree to obtain an estimate of $D$.
It is known as the `weighted' estimate of the diffusion constant.
The weighted dispersal velocity estimate compares the sum of absolute values of observed displacements to the total time, which has dimensions of a velocity.
Alternative formulations use the mean of fractions $D_b = |B|^{-1} \sum_{i\in B}d_i^2/4\Delta t_i$ or $v_b = |B|^{-1} \sum_{i\in B}d_i/\Delta t_i$ instead of the ratio of sums.
We refer to these here as `by branch' estimates.
The latter tend to be dominated by short branches and thus noisier \citep{trovao_bayesian_2015}.

The purpose of this note is to explore how sensitive phylogeographic inferences are to model assumptions and what quantities can be robustly estimated.
My focus here is only on the ability and robustness to estimate dispersal parameters and ancestral locations and I will assume perfect knowledge of the tree and the sampling locations.
So no tree-sampling will be necessary.

\section*{Consistency of summary statistics of dispersal parameters.}

The estimators in Eq.~\ref{eq:dispersal_parameters} are commonly used to summarize dispersal of lineages.
While the estimator $D_w$ directly estimates a parameter of the model and its behavior has been studied in simulations \citep{pybus_unifying_2012}, it is unclear what $v_w$ is measuring and how it behaves.
I simulated diffusion along trees from a Kingman \citep{kingman_coalescent_1982} and Yule \citep{yule_iimathematical_1997} tree ensembles and evaluated the different estimators.
Kingman trees are the classical neutral model of a constant size population, while Yule trees capture features of growing populations (see Fig.~\ref{fig:illustration_tree}).

\begin{figure}[tb]
    \includegraphics*[width=0.48\textwidth]{figures/kingman_dispersal.pdf}
    \caption{\label{fig:D_and_v}{\bf Empirical summary statistics of $D$ and ``dispersal velocity'' for different sample sizes.}
    While diffusion constant $D$ (equal to 1 in this example) can be robustly estimated from simulation data of the underlying model, the dispersal rate $v_w$ (weighted) and $v_b$ (by branch) are not well-defined quantities. Their numerical value depends strongly on sample size and there is no ground truth to compare to. The expected scaling $\sim \sqrt{n}$ of $v$ estimates is indicated by the dashed line.}
\end{figure}

Fig.~\ref{fig:D_and_v} shows empirical estimates of diffusion constants and velocity as defined in Eq.~\ref{eq:dispersal_parameters} from simulated data using freely diffusing locations along Kingman trees.
The diffusion constant $D$ can be estimated robustly from these data and the estimates are compatible with the diffusion constant used to simulate the data $D=1$.
Values of $v_w$ and $v_b$, however, depend strongly on the sample size and are incompatible with each other.
That the velocity estimates are ill-defined is not surprising: By changing the sample size, the average length of branches changes. Diffusion along each branch results in a displacement of $d_i \sim \sqrt{4D\Delta t_i}$ such that a quantity that depends on $d_i / \Delta t_i \sim 1/\Delta t_i^{1/2}$ will change with the time $\Delta t_i$ that elapsed along the branch.
With more samples in the tree, the average branch length decreases, and the velocity estimate increases.
This expected scaling $\sim \sqrt{n}$ is indicated by the dashed line in Fig.~\ref{fig:D_and_v}.
For Yule trees, the average branch length is independent of the number of samples and the velocity estimates are more stable.
But estimates for $v_w$ and $v_b$ are not consistent with each other.

@@

\section*{Effects of density regulation on population structure and dispersal}
Phylogeographic inference typically assumes that the spatial diffusion process is independent of the branching pattern of the tree and ignores coupling between the two.
Such a coupling, however, is expected since growth rates naturally depend on local population density when organisms compete for resources.
It has been known for a long time that ignoring such coupling leads to unrealistically heterogeneous population densities \citep{felsenstein_pain_1975}.
If one simulates a diffusive spatial process in a square of length $L$ with periodic boundary conditions, the population density shows strong fluctuations whenever the diffusion constant $D<L^2/T_c$, where $T_c$ is the coalescent timescale of the population.
This is seen in the examples in Fig.~\ref{fig:illustration_tree} where the population is distributed in patches corresponding to clades in the tree with large uninhabited areas.

For $D<L^2/2T_c$, the time to the most recent common ancestor $T_{MRCA}$ is too small for the population to spread evenly in the habitat.
This heterogeneity increases substantially for even lower $D$, when the population is essentially a cloud of width $\sqrt{4DT_c}\ll L$.
Fig.~\ref{fig:density_reg}A (black line) quantifies this heterogeneity as the standard deviation of population density in two-dimensional bins of different size relative to the expectation in the case of a spatially well-mixed population.

@@

Biological populations spread into areas that support them and will quickly populate fertile areas that are below their carrying capacity.
A simple model for the population density $\phi(x,t)$ would be
\begin{equation}
    \label{eq:FKPP}
    \frac{\partial \phi(x,t)}{\partial t} = D\frac{\partial^2 \phi(x,t)}{\partial x^2} + \alpha \phi(x,t)(1-\phi(x,t)/\phi_0)
\end{equation}
where $\phi_0$ is the carrying capacity and $\alpha$ is the population growth rate at zero density.
This model has a stable fixed point at $\phi=\phi_0$.
In a static environment, we thus expect the population density to settle at a constant density $\phi_0$ and density fluctuations will decay with rate $\alpha$.

While dependence of growth (and thus tree structure) on density is a key component of real-world population dynamics, it is difficult to implemented in phylogeographic inference frameworks.
It is, however, fairly straightforward to include density dependence in forward simulations and these simulations can be used to investigate the effect of such coupling between growth and spatial location on phylogeographic inferences.
We calculate the population density using a Gaussian kernel with interaction radius $r$
\begin{equation}
    \phi(\rvec) = \sum_{i\in N} \frac{e^{-|\rvec_i - \rvec|^2/2r^2}}{N\sqrt{2\pi r^2}}
\end{equation}
where the sum is over all individuals $i$ in the population of size $N$.

As soon as the interaction radius $r$ is much smaller than the habitat size, this density regulation leads to a more homogeneous population density, see Fig.~\ref{fig:density_reg}A.
But with decreasing diffusion, the population fragments into smaller and smaller subpopulations that interact only weakly and coalesce slowly.
This manifests in an increasing time to the MRCA, see Fig.~\ref{fig:density_reg}B, which at low $D$ is limited by the time the population was simulated for.
Spatial location and the coalescence process are thus strongly coupled.

Despite the substantial reduction in density fluctuations, the estimates of dispersal rates using Eq.~\ref{eq:dispersal_parameters} are not dramatically affected, see Fig.~\ref{fig:density_reg}C.
If boundary conditions are periodic, that is the habitat has the topology of a torus, estimates $D_w$ of the diffusion constant are accurate for most values of $D$ but slightly inflated when the population is heavily fragmented.
If instead of periodic boundary conditions reflective boundary conditions are used, diffusion constants are underestimated once $\sqrt{2DT_c}$ becomes comparable to the habitat size.
This underestimation of diffusion constants will then result in over-confident inference of ancestral locations, in particular for recent ancestral nodes.

These simulations reveal that both at high and at low diffusion constants, assumptions of phylogeographic inference can be problematic: if diffusion constants are much smaller than then $L^2/T_{mrca}$, the population is fragmented into effective subpopulations and this fragmentation violates the assumptions of phylodynamic models; if $D\gg L^2/T_{mrca}$, the boundaries of the habitat restrict free diffusion, resulting in underestimated dispersal parameters.

@@

In Fig.~\ref{fig:seasaw} we consider a case where the region with highest carrying capacity shifts from the left to the right in a periodic manner.
The carrying capacity is illustrated in Fig.~\ref{fig:seasaw}A for 5 time points covering half a period.
In between the two extreme left/right concentration of the carrying capacity, carrying capacity is uniform.
To avoid extinction at low dispersal rates, selection is soft, meaning that the overall population size is kept approximately constant even if the entire population is stuck in the unfavorable region.
If the dispersal rate is sufficiently high, however, the population follows the shifting carrying capacity and lineages in the tree undergo directed back-and-forth motion.
Inference of ancestral locations in such situations can be misleading as illustrated in Fig.~\ref{fig:seasaw}B.
The population was sampled at a time when it has just contracted to the left and the positions of older nodes is not estimated correctly.
Their true locations are to the right of the current population, but the inferred location are central in the shifted population.

@@

However, along with problematic inference of diffusion coefficients, estimates of ancestral locations are over-confident.
Panel Fig.~\ref{fig:seasaw}E shows $z_x^2 = \langle (\hat{x} - x)^2/\sigma_x^2 \rangle$ and $z_y^2=\langle (\hat{y} - y)^2/\sigma_y^2\rangle$ for the oldest 20\% of internal nodes, where $\hat{x}$ and $\sigma_x$ are the mean and standard deviation of the posterior of $x$, respectively (and analogously for $y$).
If the model estimated ancestral locations consistently, $z_x^2$ should be 1 on average.
Instead, it is as high as 20 for the simulations shown and this over-confident and biased estimations depend in non-monotonic ways on the true diffusion constant and the period of environmental change.
With decreasing $D$, $z_x^2$ increases largely because the expected uncertainty decreases, before peaking when the population can no longer follow the shifting carrying capacity.
A peak value of 10-20 indicates that the typical ancestral location of the deep nodes is 2-4 standard deviations away from the true location, an example of which is shown in Fig.~\ref{fig:seasaw}C.
Conversely, at high dispersal rates, the average $z_x^2$ and $z_y$ are both smaller than one since the estimated uncertainty is larger than the true uncertainty since the diffusion is constrained by the boundaries of the habitat.

>>>>

\maketitle
As organisms replicate, they also disperse in space, resulting in mixing of the population and exploration of new habitats.
The reconstruction of ancestral locations and past migrations from sampling locations of extant individuals, typically in combination with genome sequence information to infer the phylogeny, is known as phylogeography.
Phylogeographic methods are implemented in popular evolutionary analysis software such as BEAST \citep{lemey_bayesian_2009,lemey_phylogeography_2010}.

In phylogeographic inference, one can either assume that individuals migrate between discrete locations or disperse in continuous space.
Here, we consider migration in continuous space, where the underlying model is typically assumed to be diffusive, that is the probability of sampling a descendant at position $\rvec_c$ after time $t$ when the parent was at position $\rvec_p$ is given by
\begin{equation}
    P(\rvec_c| \rvec_p, t) = \frac{1}{4\pi D t}e^{\frac{|r_c - r_p|^2}{4Dt}}
\end{equation}
where $D$ is a diffusivity with dimensions $\mathrm{length}^2/\mathrm{time}$, we have assumed a two-dimensional space, and we have assumed dispersal is isotropic.
Such a diffusive process is the simplest and most natural choice in absence of directed motion or long range migration.
It is also mathematically convenient as unknown ancestral locations can be integrated in closed form and the marginal distribution of each position can be calculated exactly for a fixed tree along with the spatial likelihood.
The total likelihood of the configuration is then assumed to be the product of the spatial likelihood and the phylogenetic likelihood, that is the tree topology and branching rates of the tree are assumed to be independent of the spatial process.
This independence assumption is for example manifest in coalescent models assuming ``exchangeability'' of individuals, or birth-death models assuming identical rates on contemporary branches of the tree.

However, population dynamics are often tightly coupled to spatial location.
Changes of the environment could mean that a habitat of a population is shifting, resulting in population growth and rapid branching at the expanding edge of the habitat and/or contraction in other areas.
Similarly, invasive species or pathogens spread into areas that support high population densities but were previously unoccupied.
Examples of such shifting habitats are common and range from geological time scales, over decades (climate change), to weeks (seasonal fluctuations).
Such changes are not restricted to the physical environment, but ecological shifts can similarly affect the habitat.
In all such cases, the replication process determining the phylogeny of the sample and the spatial location are strongly coupled.

Such coupling can be accommodated in inference methods based on structured models that divide the population into discrete demes that could correspond to spatial locations that exchange migrants.
Such models are flexible enough to allow for different population dynamics in different demes \citep{vaughan_efficient_2014}.
But the extent to which violations of the independence assumption affect continuous phylogeographic inferences is not well understood.

\begin{figure}[tb]
    \includegraphics*[width=\columnwidth]{figures/illustration_tree.pdf}
    \caption{\label{fig:illustration_tree}{\bf Illustration Yule and Kingman trees with $N=100$ tips and the spatial location of the tips.}
    The total tree length of Yule trees is proportional to the number of tips $n$ and the time to the MRCA is $\log(n)$.
    Kingman trees have a total tree length that is proportional to $\log(n)$ and are characterized by rapid merging close to the present. }
\end{figure}

In a typical phylogeographic inference problem, the input data are genome sequences as well as their sampling times and sampling locations.
Software like BEAST will then sample the posterior distribution of the phylogenetic tree topology, the sampling times and locations of ancestral nodes in the tree, and model parameters like the diffusion constant $D$ and the evolutionary rate $\mu$.
The inferred quantities of interest could be estimates of the ancestral location of a particular groups of samples, or summary statistics of the dispersal process.
Here, I explore what dispersal quantities can be robustly estimated and how sensitive inferences are to the fact that coupling between population dynamics and spatial location is ignored.
My focus here is only on the ability and robustness to estimate dispersal parameters and ancestral locations and I will use simulations with perfect knowledge of the tree and the sampling locations.
So no tree-sampling will be necessary.
This simplification is justified since the tree tends to depend much more strongly on sequence information and spatial information and will allow stronger conclusions.
Furthermore, I will assume strictly diffusive dispersal without any long-range migrations.

\section*{Consistency of summary statistics of dispersal parameters.}
Before investigating the more complex issues of coupling between spatial locations and reproduction, I will first explore the properties of summary statistics in simple models where spatial location and population dynamics are independent.
Popular summaries of the dispersal process are empirical estimates of the diffusion constant $D_w$ \citep{pybus_unifying_2012,trovao_bayesian_2015} and a so-called \emph{dispersal velocity} $v_w$ \citep{dellicour_using_2017}.
These are calculated from the displacements along branches of the trees sampled from the posterior.
For each branch $i$ with parent $p_i$, the displacement $d_i = |\rvec_{p_i} - \rvec_{i}|$ and the elapsed time $\Delta t_i$ are used to calculate
\begin{equation}
    \label{eq:dispersal_parameters}
    D_w = \frac{\sum_{i\in B}d_i^2}{4\sum_{i\in B} \Delta t_i} \quad \mathrm{and}  \quad v_w = \frac{\sum_{i\in B} d_i}{\sum_{i\in B} \Delta t_i}
\end{equation}
where $B$ is the set of all branches of the tree.
Effectively $D_w$ divides the sum of all observed squared displacements by the total time elapsed on the tree to obtain an estimate of $D$.
It is known as the `weighted' estimate of the diffusion constant.
The weighted dispersal velocity estimate compares the sum of absolute values of observed displacements to the total time, which has dimensions of a velocity.
Alternative formulations use the mean of fractions $D_b = |B|^{-1} \sum_{i\in B}d_i^2/4\Delta t_i$ or $v_b = |B|^{-1} \sum_{i\in B}d_i/\Delta t_i$ instead of the ratio of sums.
We refer to these here as `by branch' estimates.
The latter tend to be dominated by short branches and thus noisier \citep{trovao_bayesian_2015}.

The estimators in Eq.~\ref{eq:dispersal_parameters} are commonly used to summarize dispersal of lineages.
While the estimator $D_w$ directly estimates a parameter of the model and its behavior has been studied in simulations \citep{pybus_unifying_2012}, it is unclear what $v_w$ is measuring and how it behaves.
I simulated diffusion along trees from a Kingman coalescent\citep{kingman_coalescent_1982} and Yule model \citep{yule_iimathematical_1925} tree ensembles and evaluated the different estimators using the known true ancestral locations.
Kingman trees are the classical neutral model of a constant size population, while Yule trees capture features of exponentially growing populations (see Fig.~\ref{fig:illustration_tree}).
Examples for each ensemble are shown in Fig.~\ref{fig:illustration_tree}.

\begin{figure*}[tb]
    \includegraphics*[width=0.99\textwidth]{figures/dispersal_stats.pdf}
    % \includegraphics*[width=0.48\textwidth]{figures/yule_dispersal.pdf}
    \caption{\label{fig:D_and_v}{\bf Empirical summary statistics of $D$ and ``dispersal velocity'' for different sample sizes.}
    While diffusion constant $D$ (equal to 1 in this example) can be robustly estimated from simulation data of the underlying model, the dispersal rate $v_w$ (weighted) and $v_b$ (by branch) are not well-defined quantities. Their numerical value depends strongly on sample size and there is no ground truth to compare to. The expected scaling $\sim \sqrt{n}$ of $v$ estimates is indicated by the dashed line.}
\end{figure*}

Fig.~\ref{fig:D_and_v} shows empirical estimates of diffusion constants and velocity as defined in Eq.~\ref{eq:dispersal_parameters} from simulated data using freely diffusing locations along Yule and Kingman trees.
The diffusion constant $D$ can be estimated robustly from these data and the estimates are compatible with the diffusion constant used to simulate the data $D=1$.
Values of $v_w$ and $v_b$ estimated from Kingman trees, however, depend strongly on the sample size and are incompatible with each other.
That the velocity estimates are ill-defined is not surprising: By changing the sample size, the average length of branches in Kingman coalescent trees changes.
Diffusion along each branch results in a displacement of $d_i \sim \sqrt{4D\Delta t_i}$ such that a quantity that depends on $d_i / \Delta t_i \sim 1/\Delta t_i^{1/2}$ will change with the time $\Delta t_i$ that elapsed along the branch.
With more samples in the tree, the average branch length decreases, and the velocity estimate increases.
This expected scaling $\sim \sqrt{n}$ is indicated by the dashed line in right panel of Fig.~\ref{fig:D_and_v}.
For Yule trees, the average branch length is independent of the number of samples and the velocity estimates are more stable.
But as for the Kingman case, the estimates for $v_w$ and $v_b$ are not consistent with each other.

@@

While the estimator $D_w$ as defined in Eq.~\ref{eq:dispersal_parameters} is well-behaved when the model is exactly diffusive, long range dispersal can affect $D_w$ dramatically.
If the probability of long jumps with distance $x<d<x+dx$ decays more slowly than $\sim dx/x^3$, the expectation value of the $D_w$ diverges.
This may mean that even $D_w$ becomes problematic when very long-tailed prior distributions for $D$ are used in relaxed random walk models.

\section*{Effects of density regulation on population structure and dispersal}
Phylogeographic inference typically assumes that the spatial diffusion process is independent of the branching pattern of the tree and ignores coupling between the two.
Such a coupling, however, is expected since organisms compete for resources and growth rates naturally depend on local population density.
It has been known for a long time that ignoring such coupling leads to unrealistically heterogeneous population densities \citep{felsenstein_pain_1975}, as is evident in the example of the Kingman tree in Fig.~\ref{fig:illustration_tree}.
If one simulates a diffusive spatial process in a square of length $L$ with periodic boundary conditions, the population density shows strong fluctuations whenever the diffusion constant $D<L^2/T_c$, where $T_c$ is the coalescent timescale of the population.
Fig.~\ref{fig:density_reg}A (black line) quantifies this heterogeneity as the standard deviation of population density in two-dimensional bins of size $L/5$ relative to the expectation in the case of a spatially well-mixed population when each individual picks its location at random.
For $D<L^2/2T_c$, the time to the most recent common ancestor $T_{mrca}$ is too small for the population to spread evenly in the habitat.
This heterogeneity increases substantially for even lower $D$, when the population is essentially a cloud of width $\sqrt{4DT_c}\ll L$.

@@

Biological populations spread into areas that support them and will quickly populate fertile areas that are below their carrying capacity.
A simple model for the population density $\phi(\rvec,t)$ would be
\begin{equation}
    \label{eq:FKPP}
    \frac{\partial \phi(\rvec,t)}{\partial t} = D \Delta_{\rvec} \phi(\rvec,t) + \alpha \phi(\rvec,t)(1-\phi(\rvec,t)/\phi_0)
\end{equation}
where $ \Delta_{\rvec}$ is the two-dimensional Laplace operator, $\phi_0$ is the carrying capacity, and $\alpha$ is the population growth rate at zero density.
Such logistic density regulation was proposed by \citet{bolker_using_1997} and further analyzed by \citet{etheridge_survival_2004}.
The density has a stable fixed point at $\phi=\phi_0$ we thus expect the population density to settle at a constant density $\phi_0$ and density fluctuations will decay with rate $\alpha$.

While the dependence of growth (and thus tree structure) on density is a key component of real-world population dynamics, it is difficult to implement in phylogeographic inference frameworks.
It is, however, fairly straightforward to include density dependence in forward simulations and these simulations can be used to investigate the effect of such coupling between growth and spatial location on phylogeographic inferences.

Instead of the coalescent trees used above, I now used an individual based simulation forward in time where individuals are assigned a fitness $f=\alpha(1-\hat{\phi}/\phi_0) + (1-N/N_0)$ and produce a Poisson distributed number of offspring with mean $1+f$.
The first term determines the fitness due to the local population density and has the tendency to stabilize this density at $\phi_0$, the second term $(1-N/N_0)$ keeps the overall population close to the target population size $N_0$.
The population density is calculated using a Gaussian kernel with interaction radius $r$
\begin{equation}
    \ref{eq:density_estimator}
    \hat{\phi}(\rvec, t) = \sum_{i\in N} \frac{e^{-|\rvec_i - \rvec|^2/2r^2}}{\sqrt{2\pi r^2}}
\end{equation}
where the sum is over all individuals $i$ in the population of size $N$ and $\rvec_i$ is location of individual $i$ at time $t$.
The spatial locations of the offspring are chosen from a 2-D Gaussian centered on the parent with a variance to match the diffusion constant.
The simulation records the tree of population, including their ancestral location and times.
In addition to the true location of ancestral nodes, I also infer their locations from the locations of the extant samples.
Since the model is purely diffusive, the marginal distributions of ancestral locations and their mean and variance can be calculated exactly using recursion relations.

The radius $r$ of the density estimate $\hat{\phi}(\rvec, t)$ in Eq.~\ref{eq:density_estimator} determines the distance over which different individuals compete for resources.
As soon as $r$ is much smaller than the habitat size, this density regulation leads to a more homogeneous population density, see Fig.~\ref{fig:density_reg}A.
But with decreasing diffusion, the population fragments into smaller and smaller subpopulations that interact only weakly and coalesce slowly.
This manifests in an increasing time to the MRCA, see Fig.~\ref{fig:density_reg}B, which at low $D$ is limited by the time the population was simulated for.
Spatial location and the coalescence process are thus strongly coupled.

Despite the substantial reduction in density fluctuations, the estimates of dispersal rates using Eq.~\ref{eq:dispersal_parameters} are not dramatically affected by density regulation, see Fig.~\ref{fig:density_reg}C.
If boundary conditions are periodic, that is the habitat has the topology of a torus, estimates $D_w$ of the diffusion constant are accurate for most values of $D$ but slightly inflated when the population is heavily fragmented.
If instead of periodic boundary conditions reflective boundary conditions are used, diffusion constants are underestimated once $\sqrt{2DT_c}$ becomes comparable to the habitat size.
This underestimation of diffusion constants will then result in over-confident inference of ancestral locations, in particular for recent ancestral nodes.

These simulations reveal that both at high and at low diffusion constants, assumptions of phylogeographic inference can be problematic: if diffusion constants are much smaller than then $L^2/T_{mrca}$, the population is fragmented into effective subpopulations and this fragmentation violates the assumptions of phylodynamic models.
A structured model would be more appropriate in this case.
If $D\gg L^2/T_{mrca}$, the boundaries of the habitat restrict free diffusion, resulting in underestimated dispersal parameters, while estimates of deeper ancestral locations have uncertainties comparable to the habitat size.

@@

In Fig.~\ref{fig:seasaw} we consider a case where the region with highest carrying capacity shifts from the left to the right in a periodic manner.
The carrying capacity is illustrated in Fig.~\ref{fig:seasaw}A for 5 time points covering half a period.
In between the two extreme left/right concentration of the carrying capacity, carrying capacity is uniform.
To avoid extinction at low dispersal rates, selection is soft, meaning that the overall population size is kept approximately constant even if the entire population is stuck in the unfavorable region.
If the dispersal rate is sufficiently high, however, the population follows the shifting carrying capacity and lineages in the tree undergo directed back-and-forth motion.
Inference of ancestral locations in such situations can be misleading as illustrated in Fig.~\ref{fig:seasaw}B.
The population was sampled at a time when it had just contracted to the left and the positions of older nodes are not estimated correctly.
Their true locations are to the right of the current population, but the inferred location are central in the shifted population.
This is expected as there is no information in the current sample to suggest otherwise, but it highlights the potential to mislead.

@@

However, along with problematic inference of diffusion coefficients, estimates of ancestral locations are over-confident.
Panel Fig.~\ref{fig:seasaw}E shows $z_x^2 = \langle (\hat{x} - x)^2/\sigma_x^2 \rangle$ and $z_y^2=\langle (\hat{y} - y)^2/\sigma_y^2\rangle$ for the oldest 20\% of internal nodes, where $\hat{x}$ and $\sigma_x$ are the mean and standard deviation of the posterior of $x$, respectively (and analogously for $y$).
If the model estimated ancestral locations consistently, $z_x^2$ should be 1 on average.
Instead, it is as high as 20 for the simulations shown and these over-confident and biased estimates depend in non-monotonic ways on the true diffusion constant and the period of environmental change.
With decreasing $D$, $z_x^2$ increases largely because the expected uncertainty decreases, before peaking when the population can no longer follow the shifting carrying capacity.
A peak value of 10-20 indicates that the typical ancestral location of the deep nodes is 2-4 standard deviations away from the true location, an example of which is shown in Fig.~\ref{fig:seasaw}C.
Conversely, at high dispersal rates, the average $z_x^2$ and $z_y$ are both smaller than one since the estimated uncertainty is larger than the true uncertainty since the diffusion is constrained by the boundaries of the habitat.
